---
title: "Lecture 5. Missing data"
author: "Sabrina Nardin"
output: html_document
date: "2024-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load libraries and data

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
```

```{r}
data("penguins")
```


## Missing data: definitions

Our book distinguishes missing data between: 
* explicit: cells where you see a "NA" 
* implicit: absent data (e.g., an entire row is absent because not collected, etc.) 

In this tutorial, we focus on explicit missing data. I recommend reviewing [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science" for implicit data.

Explicit missing data are "NA" (Not Available) or "NaN" (Not a Number). Both "NA" and "NaN" behave similarly in many contexts, though "NaN" specifically indicates issues arising from computations. "NaN" generally results from mathematical operations with indeterminate values (e.g., `0/0` or `0 * Inf`). 

Missing data behavior: any operation involving an unknown value will be unknown. See our book [Chapter 12.2.2 Missing values](https://r4ds.hadley.nz/logicals#sec-na-comparison) for more. 
 
```{r}
NA > 5
```

```{r}
sum(3 + 1 + 4 + NA)
```

```{r}
mean(c(3,1,4))
mean(c(3,1,4,NA))
```


## handle missing data in R

Most common ways to handle missing data: 

* `is.na()` 
* `na_rm = TRUE`
* `drop_na()`


`is.na()` to identify missing data (detects both "NA" and "NaN"); returns TRUE for missing values and FALSE for anything else

```{r}
is.na(c(3, 4, NA, 12))
```

```{r}
# do not display this output
is.na(penguins$body_mass_g)
```

```{r}
# much better
sum(is.na(penguins$body_mass_g))
```

```{r}
# filter and missing data
filter(penguins, is.na(sex))
filter(penguins, !is.na(sex))

# wrong code 
# filter(penguins, sex == NA)
```


`na_rm = TRUE` to ignore missing values when performing a given operation; typically used with `summarize()` when calculating things like mean, sum, standard deviation. Important: here missing values are not dropped, they are just ignored for that given calculation

```{r}
# summarize and missing data
penguins %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
```


`drop_na()` to remove observations (rows) with missing values, in all columns or in specific ones. 
```{r}
penguins %>%
  drop_na(body_mass_g) %>%
  summarize(avg_mass = mean(body_mass_g))

# check what happens if you do not drop them
```

```{r}
penguins %>%
  drop_na(body_mass_g) %>% 
  filter(species == "Adelie") 
```

In all these cases, remember to save your output in a new dataframe (or "df" for short)


## Ways to fill or replace missing data in R

See [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science".

Main functions: 
* `replace_na()` to replace missing data with some other value
* `fill()` (from `tidyr`)
* `coalesce()`


## Factors and empty groups

Example taken from [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science"

Data of health info about people
```{r}
health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)
health
```

Count number of smokers
```{r}
health |> count(smoker)
```

Keep all groups, even those not observed in the data (the five people in this data are all non-smokers)
```{r}
health |> count(smoker, .drop = FALSE)
```

This is useful for plotting. Observe the difference between these two bar charts: 
```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete()
```

```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```
