---
title: "Lecture 5. Rename and recode variables"
author: "Sabrina Nardin"
output: html_document
date: "2024-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
```

```{r}
data("penguins")
```


## Explore the data

```{r}
glimpse(penguins)
```

```{r}
summary(penguins)
```


## Renaming and recoding variables


### Renaming variables

To change variable names the most common method is `rename()` with `new_name = old_name` syntax.For example, we want to rename into `specie` the current variable `species`: 

```{r}
# check before renaming
penguins %>% select(species)
```

```{r}
# rename
penguins %>% 
  rename(specie = species)  
```

```{r}
# remember to save results in a new df if you want to save it
p <- penguins %>% 
  rename(specie = species) 
p %>% select(specie)
```


### Recoding variables

There are a few ways to change the name of the levels of categorical variables, we review two of them. For example, we want to rename the levels of the categorical variable `sex` into 1 for male, and 0 for female.

Option 1: use `mutate()` + `recode()` with `old_name = new_name` syntax. See: https://dplyr.tidyverse.org/reference/recode.html
```{r}
# check before recoding
count(penguins, sex)

# equivalent code
#penguins %>% count(sex)
```

```{r}
# mutate + recode
penguins %>%
  mutate(sex = recode(
    sex,
    male = 1,   # old = new
    female = 0)) %>%  
  count(sex)
```


Option 2: another popular method to change the name of levels of categorical variables is `mutate()` + `case_when()` with `old_name = new_name` syntax. See: https://dplyr.tidyverse.org/reference/case_when.html
```{r}
# mutate + case_when
penguins %>%
  mutate(sex = case_when(
    sex == "female" ~ 0,  # old ~ new
    sex == "male" ~ 1
  )) %>%
  count(sex)

# note the use of "" around female and male (vs. above code using recode
```

Notice we always use `mutate()` when we need to manipulate variables! `mutate()` creates new columns from existing ones or modify existing columns. For example, we used it here together with functions for renaming or recording variables. And we also used it when we needed to change a character variable into a factor (see examples in lecture 5 slides).


## Syntactic and non-syntactic variable names

### Syntactic names

A syntactic name is what R considers a valid name: letters, digits, `.` and `_` but it can’t begin with symbols or a with a digit.

```{r}
count(penguins, bill_length_mm)
```

Here `bill_length_mm` is a column of the `penguins` data frame. It has a syntactic name, so we can refer to it directly. 

Examples of syntactic or valid variable names:
```
bill_length_mm
bill.length.mm
bill.length_mm    # bad coding style but "valid" name
ideology
ideology_7_point
id
```

### Non-syntactic names

A non-syntactic name is a name that R does not consider a valid name: names that contain spaces, start with a digit or a symbol, or use reserved words such as TRUE, NULL, if, or function names. See the complete list of reserved names by typing `?Reserved` in your Console.

Example of non-syntactic or invalid variable names:
```
bill length mm
bill_ length_mm
social conservative
social_ conservative
social-conservative
7-point ideology
_id
if
```

### Why should we care? Use backticks when you have a non-syntactic name!

You do not want to deliberately create non-syntactic names, BUT you need to understand how they work because you’ll come across them when working with social sciences data, especially if the data that has been created outside R. This happens more frequently that you would expect!

To refer to non-syntactic names: (1) use backticks, not quotes; or/and (2) recode them.

Example: imagine you have loaded into R data on ideology by country, assembled by someone else in Excel. As you start exporing the data, you find a variable labeled as follows:

```
7-point ideology <- c("communism", "anarchism", "fascism")
```

To use it without changing it, you must use backticks, otherwise R will throw an error:
```{r}
`7-point ideology` <- c("communism", "anarchism", "fascism")
```


