---
title: "Improve the bar chart from Lecture 4 using `scorecard` data"
output: html_document
author: Sabrina Nardin
---

Load libraries and data
```{r, message = FALSE}
library(tidyverse)
library(rcis)
data(scorecard)
```

Check
```{r}
glimpse(scorecard)
```

This is the code to reproduce the same plot from TASK 4 of Lecture 4:
```{r barplot_cat}
ggplot(scorecard,
  mapping = aes( x = type, fill = state)) +
  geom_bar(position = "stack")
```

This (stacked) bar chart represents the total number of schools in the U.S. by school type (n = 3) and by state (n = 54). We have two categorical variables: `type` and `state`.

Each bar represents one category of a categorical variable, and is divided into segments representing the categories of the other categorical variable. But although it is not wrong, this is not a good graph: it is visually crowded and as such is not as informative as it could be.

Does swapping the x and y axes and adjusting the legend (by rotating the labels) result in a more effective bar chart? Notice I am also adding titles and axis labels for enhanced clarity (not necessary in truly exploratory graphs!):
```{r barplot_cat_2}
ggplot(scorecard,
  mapping = aes(x = state,
                fill = type)) +
  geom_bar(position = "stack") +
  labs(
    title = "Number of Schools by State and Type",
    x = "State",
    y = "Count",
    fill = "School Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.5))
```

This is a bit better, but we’re still not quite there! The point is that there’s only so much we can do with `ggplot2`: if the problem lies in the input data, we need to use `dplyr` to manipulate the variables *before* visualizing them!

For example, I would aggregate the 54 states using some meaningful criteria (decided by the researcher) and then display them. 

**Task: group current U.S. states (variable `state`) into four U.S. regions (create a new variable `region`) and label the regions as "Northeast", "Midwest", "South", "West". Then create a bar chart using the regions (vs. states) and school types**

Hint: You might want to define categories manually, using `mutate()` and `case_when()`, then use other dplyr functions to summarize the aggregated number of schools by region

```{r}
# add as many code chuncks as you'd like
```

```{r}

```

