---
title: "Missing data"
output: html_document
author: Sabrina Nardin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code chunk

* PC: Cntr + Alt + I
* Mac: Cmd + Option + I

Pipe

* PC: Ctrl + Shift + M 
* Mac: Cmd + Shift + M

```{r}
library(tidyverse)
library(palmerpenguins)
```


```{r}
data("penguins")
```

## Missing data: definitions

Our book distinguishes missing data between: 
* explicit: cells where you see a NA 
* implicit: absent data (e.g., an entire row is absent because not collected, etc.) 

In this tutorial, we focus on explicit missing data. I recommend [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science" for implicit data.

Explicit missing data are "NA" (Not Available) or "NaN" (Not a Number).
"NaN" generally results from mathematical operations with indeterminate values (e.g., `0/0` or `0 * Inf`). Both "NA" and "NaN" behave similarly in many contexts, though "NaN" specifically indicates issues arising from computations.

Missing data ("NA") behavior: any operation involving an unknown value will be unknown. See our book [Chapter 12.2.2 Missing values](https://r4ds.hadley.nz/logicals#sec-na-comparison) for more. 
 
```{r}
NA > 5
```

```{r}
sum(3 + 1 + 4 + NA)
```

```{r}
# mean(3+1+4+NA) won't work: what is inside mean() is first evaluated giving 8, thus mean(8) = 8

mean(c(3,1,4))
mean(c(3,1,4,NA))
```


## Ways to handle missing data in R

Most common: 
* `is.na()` 
* `na_rm = TRUE`
* `drop_na()`


`is.na()` to identify missing data (detects both "NA" and "NaN"); returns TRUE for missing values and FALSE for anything else

```{r}
is.na(c(3, 4, NA, 12))
```

```{r}
is.na(penguins$body_mass_g)
```

```{r}
sum(is.na(penguins$body_mass_g))
```

```{r}
# filter and missing data
filter(penguins, is.na(sex))
filter(penguins, !is.na(sex))

# wrong code 
# filter(penguins, sex == NA)
```


`na_rm = TRUE` to ignore missing values when performing a given operation; typically used with `summarize()` when calculating things like mean, sum, standard deviation (important: missing values are not dropped, they are just ignored for that given calculation)

```{r}
# summarize and missing data
penguins %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))
```


`drop_na()` to remove observations (rows) with missing values, in all columns or in specific ones
```{r}
penguins %>%
  drop_na(body_mass_g) %>%
  summarize(avg_mass = mean(body_mass_g))

# check what happens if you do not drop them
```

```{r}
penguins %>%
  drop_na(body_mass_g) %>% 
  filter(species == "Adelie") 
```


In all these cases, remember to save in a new dataframe (or "df" for short)!


## Ways to fill or replace missing data in R

See [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science".

Main functions: 
* `replace_na()` to replace missing data with some other value
* `fill()` (from `tidyr`)
* `coalesce()`


## Factors and empty groups

Example taken from [Chapter 18](https://r4ds.hadley.nz/missing-values) of "R for Data Science"

Data of health info about people
```{r}
health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)
```

Count number of smokers
```{r}
health |> count(smoker)
```
Keep all groups, even those not observed in the data (the five people in this data are all non-smokers)
```{r}
health |> count(smoker, .drop = FALSE)
```

This is useful for plotting. Observe these two bar charts: 

```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete()
```

```{r}
ggplot(health, aes(x = smoker)) +
  geom_bar() +
  scale_x_discrete(drop = FALSE)
```
