---
title: "More dplyr"
output: html_document
author: Sabrina Nardin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code chunk

* PC: Cntr + Alt + I
* Mac: Cmd + Option + I

Pipe

* PC: Ctrl + Shift + M 
* Mac: Cmd + Shift + M

```{r}
library(tidyverse)
library(palmerpenguins)
```


```{r}
data("penguins")
```


## Explore the data

```{r}
# head and tail
head(penguins)
tail(penguins)
rbind(head(penguins, 5), tail(penguins, 5))
glimpse(penguins)
```

```{r}
# dataset summaries
str(penguins)
summary(penguins)
```


## Basic dplyr verbs

### Selecting columns, filtering by row, reordering (both columns and rows)

`select()` extracts variables based on their names (operates on columns)
```{r}
# select two specific variables using their names
select(penguins, species, body_mass_g)

# select all variables except two using the minus sign
select(penguins, -year, -island)

# select a range of contiguous/adjacent variables
select(penguins, species:bill_length_mm)

# select all variables except a range of contiguous ones
select(penguins, -(species:island))

# starts with
select(penguins, starts_with("bill"))

# ends with
select(penguins, ends_with("mm"))
```


`filter()` extracts observations based on their values (operates on rows); logical operators are commonly used with it 
```{r}
# ways to check variable values 
levels(penguins$species) 
unique(penguins$species)
count(penguins, species)

# filter by island and species
filter(penguins, (island == "Biscoe" & species == "Gentoo"))
filter(penguins, island == "Biscoe", species == "Gentoo")

# filter by multiple conditions
filter(penguins, species %in% c("Adelie", "Gentoo"), island != "Biscoe")
```


`select()` + `filter()`
```{r}
penguins %>% 
  select(species, sex, year, island) %>% 
  filter(island == "Torgersen")
```


`arrange()` changes the order of observations based on their values (e.g., changes rows order)
```{r}
penguins %>%
  select(species, contains("bill")) %>%
  filter(!species == "Adelie") %>% 
  arrange(desc(bill_length_mm)) 

penguins %>%
  select(species, contains("bill")) %>%
  filter(!species == "Adelie") %>% 
  arrange(desc(species), bill_length_mm)
```


`relocate()`: changes the order of variables based on their name (e.g., changes columns order)
this is new (have not learned it yet, not in last week's slides)
```{r}
penguins %>%
  relocate(year, sex)

penguins %>% 
  relocate(year, .after = island)
```


### Manipulating variables and doing operations with them

`mutate()` creates new columns from existing ones or modify existing ones; it does not reduce the dimensions of the original data; https://dplyr.tidyverse.org/reference/mutate.html
```{r}
penguins %>% 
  mutate(bill_ratio = bill_length_mm / bill_depth_mm)

# remember to save data into the df! dply functions do not modify variables in-place!
```


`summarize()` collapses the data to a smaller number of rows to summarize the original data; you can use summarize to create all sort of summary statistics; https://dplyr.tidyverse.org/reference/summarise.html
```{r}
penguins %>% 
  summarize(body_mass_mean = mean(body_mass_g, na.rm = TRUE),
            body_mass_sd = sd(body_mass_g, na.rm = TRUE),
            n = n())

penguins %>% 
  filter(species == "Gentoo") %>% 
  summarize(body_mass_mean = mean(body_mass_g, na.rm = TRUE),
            body_mass_sd = sd(body_mass_g, na.rm = TRUE),
            body_mass_count = n())
```


`group_by()` aggregates categorical variables into their groups, often used together with summarize, because alone is not that useful (you want to group by and then calculate some summary statistics for each group); technically, you can use group_by with any type of variable (also numerical) but often it makes sense only for categorical variables
```{r}
# group by alone? not that useful
penguins %>% 
  group_by(island, sex)

# group by and count
penguins %>% 
  group_by(island, sex) %>% 
  #drop_na(sex) %>%
  count()

#  group by and calculate mean body mass
penguins %>% 
  group_by(island, sex) %>%
  drop_na(sex) %>%
  summarize(body_mass_mean = mean(body_mass_g))
```


`count()` counts total observations by group
```{r}
# total penguins by sex without count
penguins %>% 
  group_by(sex) %>% 
  summarize(total = n())

# with count
penguins %>% 
  count(sex)
```


New verbs (that we did not mention last week):

`n_distinct()` counts the number of unique values in a given column, used together with summarize
```{r}
penguins %>%
  summarise(num_species = n_distinct(species))
```


`distinct()` removes duplicate rows https://dplyr.tidyverse.org/reference/distinct.html
```{r}
penguins %>%
  distinct()

penguins %>%
  distinct(species, island)

# remove duplicate rows based on variable species
penguins %>%
  distinct(species, .keep_all = TRUE)
```
Other than dplyr functions, R has "base" functions to deal with duplicates elements such as `duplicated()` for identifying them and `unique()` for extracting them.


`across()` eliminates redundant syntax for column-wise operations, because allows to perform the same operation to multiple columns simultaneously; common use: inside functions like `summarize()` or `mutate()`

Examples with and without `across()` with numerical variables
```{r}
# mean for a bunch of variables 
penguins %>%
  summarize(
    mean_bill_length_mm = mean(bill_length_mm, na.rm = TRUE),
    mean_bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE),
    mean_flipper_length_mm = mean(flipper_length_mm, na.rm = TRUE),
    mean_body_mass_g = mean(body_mass_g, na.rm = TRUE)
  )

# same operation with across
penguins %>%
  summarize(
    across(
      c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g), 
      ~ mean(.x, na.rm = TRUE)  # function(x) mean(x, na.rm = TRUE)
    ))
```

Examples with and without `across()` with categorical variables
```{r}
# number of distinct entries for columns without across
penguins %>%
  summarise(distinct_species = n_distinct(species),
            distinct_island = n_distinct(island),
            distinct_year = n_distinct(year))

# number of distinct entries for columns with across
penguins %>%
  summarise(across(c(species, island, year), 
                   n_distinct))

# the opposite
penguins %>%
  summarise(across(-c(species, island, year), 
                   n_distinct))

# everything
penguins %>%
  summarise(across(everything(), 
                   n_distinct))

# everything with sum of na
penguins %>%
  summarise(across(everything(), 
                   ~sum(is.na(.))
                   ))
```